Experience Report: Haskell in Computational Biology

### 2012-06-28 mrfy_experience_report.tex
s hat.
splice more things into code

In a hidden Markov model, each transition is actually labeled with a
score, not a probability.
And we include the labels of the states that are
connected:\remark{\emph{Why} are the labels there?
And since they are apparently never used, why do they have strictness annotations???}
\begin{smallverbatim}
data TProb =
     TProb { logProbability :: Score
           , fromState :: StateLabel
           , toState :: StateLabel
           }
\end{smallverbatim}
In~a Plan7 hidden Markov model, as discussed above, 
there are seven transitions into each node:
\begin{smallverbatim}
data TProbs =
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb
            , d_m :: TProb, d_d :: TProb }
\end{smallverbatim}
In~our code, we select from this 7-tuple using the \texttt{ascore}
function:\remark{Norman: please put ascore into a file}
\smallverbatiminput{ascore}
In MRFy, the first two arguments to \texttt{ascore} are known at compile
time, and we expect \texttt{ascore} to be inlined.

@@

To~account for beta strands, computational biologists must use both a
richer model of protein structure and a revised version of Viterbi's
equations.
The~richer model recognizes that only certain pairs of amino acids are
able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
depends on~$x_{\pairedwith j}$.
The~relevant term in
Viterbi's equations 
$V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
Columns~$j$ and $\pairedwith j$ can be as close together as a few
columns or as far apart as a few hundreds of columns.

The presence of beta strands also changes the model.
Within a beta strand, amino acids are not inserted or deleted;
a~bonded pair of beta strands is modeled by
 a pair of sequences of match states.
Between the beta strands, the model is structured as before.
The~combined model, an~example of which is shown in \figref{mrf}, is called a
\textit{Markov random field}. 
Becayuse $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
also on the nonlocal $x_{\pairedwith j}$, 
dynamic programming cannot compute the maximum likelihood quickly
\cite{Menke:2010ti,Daniels:2012}.
MRFy explores new methods of \emph{approximating}
maximum likelihood.

>>>>

In a hidden Markov model, each transition is actually labeled with a
score, not a probability.
And we include the labels of the states that are
connected:\remark{\emph{Why} are the labels there?
And since they are apparently never used, why do they have strictness annotations???}
\smallverbatiminput{tprob}
In~a Plan7 hidden Markov model, as discussed above, 
there are seven transitions into each node:
\smallverbatiminput{tprobs}
In~our code, we select from this 7-tuple using the \texttt{ascore}
function:\remark{Norman: please put ascore into a file}
%\smallverbatiminput{ascore}
In MRFy, the first two arguments to \texttt{ascore} are known at compile
time, and we expect \texttt{ascore} to be inlined.

@@

To~account for beta strands, computational biologists must use both a
richer model of protein structure and a revised version of Viterbi's
equations.
The~richer model recognizes that only certain pairs of amino acids are
able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
depends on~$x_{\pairedwith j}$.
The~relevant term in
Viterbi's equations 
$V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
Columns~$j$ and $\pairedwith j$ can be as close together as a few
columns or as far apart as a few hundreds of columns.
Because $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
also on the nonlocal $x_{\pairedwith j}$, 
dynamic programming cannot compute the maximum likelihood quickly
\cite{Menke:2010ti,Daniels:2012}.
MRFy explores new methods of \emph{approximating}
maximum likelihood.

The presence of beta strands also changes the model.
Within a beta strand, amino acids are not inserted or deleted;
a~bonded pair of beta strands is modeled by
 a pair of sequences of match states.
Between the beta strands, the model is structured as before.
The~combined model, an~example of which is shown in \figref{mrf}, is called a
\textit{Markov random field}.
